---
title: "PSY 8814 - Lab 03: Loops and Set Theory"
author: "Linh Nguyen"
date: "2023-09-22"
output: 
  pdf_document:
    toc: TRUE
urlcolor: blue
---

\hfill\break
  
You can find the interactive webpage to edit and/or run the code yourself [here](https://nguyenllpsych.quarto.pub/psy8814-lab03/). Please wait for WEBR STATUS to be Ready.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.height = 4)
```

```{r libraries}
library(dplyr)    # for wrangling and pipe
library(ggplot2)  # for plotting
```

# I. Loops

## 1. Reproducible random state

We have previously seen the `set.seed()` function in other lab notes. This function is specifically useful when we are doing anything *random* in R, such as random sampling from `rnorm()` or `sample()`. Computers are never truly random but are instead pseudo-random. This means that you can recreate a random event later on by setting the random *seed*. This is helpful for reproducibility. It ensures that you will all get the same results on your individual machines, and that I will get the same result whenever I rerun the code chunk. You can either set the seed at the beginning of a script file, beginning of a code chunk, or right before a random process.

```{r seed}
# Set seed
set.seed(12345)

# Randomize 1 through 10
sample(x = 1:10, size = 10, replace = FALSE)
```

## 2. For-Loops

Whenever you are repeating a process multiple times, you can use for-loops to simplify your code. 


### Looping through numbers

For example, you might want to simulate 100 vectors of random numbers, and for each vector compute the mean. I will then plot the histogram of all these 100 means.

```{r loop-num-setup}
# Set seed for reproducibility
set.seed(8814)

# number of trials
n_trials <- 100

# create an empty vector of means to be populated during the loops
# numeric(n) creates a vector of n zeroes
mean_vector <- numeric(n_trials)
```

```{r loop-num}
# start for-loop
# 1:n_trials is a sequence starting at 1 and ending at n_trials (100 here)
for(i in 1:n_trials) {
  
  # draw 100000 random trials from 
  # a uniform distribution
  y <- runif(n = 100000, min = 0, max = 1)
  
  # Compute and save mean of y at position i in mean_vector
  mean_vector[i] <- mean(y)
}
```

```{r loop-num-plot}
# view the first few values of vector
mean_vector[1:10]

# plot the vector
hist(mean_vector, 
     xlab = "Sampled Means",
     main = "Histogram of 100 sampled means from uniform distributions")
```

### Looping through values

You can loop through anything! You can loop through a vector of variable names, a list of data frames, a list of plots, etc.

Here, I am showing an example of looping through a vector of variable names:

```{r loop-char-setup}
# remove scientific notation
options(scipen = 999)

# load state dataset from the built-in datasets package
data(state)
state <- state.x77[, 3:6]
head(state) %>%
  knitr::kable()

# create a vector of variable names of the state dataset
name_vector <- variable.names(state)
name_vector
```

```{r loop-char}
# create a for-loop to iterate through name_vector
for (name in name_vector) {
  
  # first print out the name to double check
  print(paste("The current variable is", name))
  
  # then create a summary of the variable in the state dataset
  print(summary(state[, name]))
  
  # then create a histogram of the variable in the state dataset
  hist(state[, name],
       main = paste("Histogram of", name),
       xlab = name)
}
```

### Try it yourself!

Draw 150 samples each with 5000 observations from a normal distribution with mean $\mu = 100$ and standard deviation $\sigma = 15$. For each sample, calculate and store the mean in a vector. At the end, take the mean of all the sample means and the standard deviation of all the sample means. Print these 2 results.

```{r loop-practice}
# insert your code here!

```


# II. Set Theory

R has built-in functions for set theory operations. To demonstrate these functions, let's first create two sets, $A$ and $B$.

```{r set-theory}
# Let's create some sets!
(A <- c("red", "orange", "green", "blue", "plum"))
(B <- c("apple", "orange", "plum", "melon", "kiwi")) 
```

## 1. Union

The `union()` function finds the union of the two sets, $A\cup B$, with all elements that are in either vector $A$ or $B$.

```{r union}
# Compute the union
union(A, B)
```

## 2. Intersect

Similarly, `intersect()` finds the intersection, $A\cap B$, which is all elements that are in *both* $A$ and $B$.

```{r intersect}
# Compute the intersection
intersect(A, B)
```

Here we see that "orange" and "plum" are both colors and fruits!

## 3. Set difference and complement

The `setdiff()` function finds the set difference $A \setminus B$: elements of $A$ that are not also in $B$. 

```{r setdiff}
# Compute the difference of these sets
# A - B =/= B - A
setdiff(A, B)
setdiff(B, A)
```

## 4. Inclusion

We can use `is.element()` or `%in%` to check whether a single element belongs to a set. Here, we're asking R whether "apple" belong to our sets $A$ and $B$.

```{r inclusion}
# is apple in the color set?
is.element("apple", A)

# is apple in the fruit set?
"apple" %in% B
```


# III. References

- This document is adapted from materials from Justin Kracht and Allie Cooperman.
